---
title: "RAB741_report"
author: "Marion Hardy"
date: "2023-02-02"
output: 
  html_document:
    toc: true 
    theme: spacelab 
    highlight: monochrome
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, cache = TRUE, echo = FALSE, warning = F, cache.lazy = F)
knitr::opts_chunk$set(fig.width=10, fig.height=15) 

library(tidyverse)
library(ggrepel)
library(clusterProfiler)
library(msigdbr)

```

```{r Data loading, include=FALSE}

res_tbl <- read_csv("./data_output/sel_vs_ctrl_T96/res_tbl.csv")
diff <- read_csv("./data_output/sel_vs_ctrl_T96/sign_different.csv")
res_tbl1 <-  res_tbl %>% filter(!is.na(ENTREZID), padj<0.05)

```

# Introduction

Data from Raphaële's samples

-   Timepoints: 0, 48 and 96 hours

-   Conditions: w/ or w/o tryptophan (Trp or selected)

-   Technical duplicates

For this report, we are only comparing gene expression between Trp negative and control at timepoint 96h.

When something appears as "enriched", it can be interpreted as enriched in cells deprived of tryptophan.

# Volcano plot for T96

```{r, fig.height = 12, fig.width = 15}

res_tbl %>%
  filter(!is.na(padj)) %>%
  ggplot(aes(x = log2FoldChange, y = -log10(padj),
             color = padj < 0.05 & abs(log2FoldChange) > 1, 
             label = ifelse(padj<0.05&log2FoldChange>=1|padj<0.05&log2FoldChange<=-1,as.character(gene),''))) +
  scale_colour_manual(values = c("gray", "firebrick3")) +
  geom_point(size = 0.5) +
  geom_hline(yintercept = -log10(0.05)) +
  geom_vline(xintercept = 1) +
  geom_vline(xintercept = -1)+
  geom_text_repel(max.overlaps = 15)+
  labs(title = "Selected vs control at T96")+
  theme_bw()

```

# GSEA analysis

I tested the Gene Set Enrichment Analysis on three differently ranked list of our significantly differently expressed genes. Ranking based on:

-   Log fold change (normalized enrichment score can be negative)

-   Adjusted pval (normalized enrichment score can not be negative)

-   Test statistic (normalized enrichment score can not be negative)

## LogFC ranking

```{r}
ordered_genes_fc <- res_tbl1$log2FoldChange
names(ordered_genes_fc) <- res_tbl1$gene
ordered_genes_fc <- sort(ordered_genes_fc, decreasing = T)

set.seed(054057)

```

### Gene ontology

```{r}

mm_GO_sets <- msigdbr(
  species = "Mus musculus", 
  category = "C5") # for gene ontology collection

gsea_results_go <- GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH", 
  TERM2GENE = dplyr::select(
    mm_GO_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_go, x = "NES", showCategory = 30)+ ggtitle("GSEA LFC Gene ontology")

gsea_results_go %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_GO_sets")

```

### Hallmark

```{r, fig.height = 10, fig.width = 8}

mm_hallmark_sets <- msigdbr(
  species = "Mus musculus", 
  category = "H") # for hallmark collection

gsea_results_hall <- GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH", 
  TERM2GENE = dplyr::select(
    mm_hallmark_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_hall, x = "NES", showCategory = 30)+ ggtitle("GSEA LFC Hallmarks")

gsea_results_hall %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_hallmark_sets")
```

### Reactome

```{r}

mm_reactome_sets <- msigdbr(
  species = "Mus musculus", 
  category = "C2",
  subcategory = "CP:REACTOME") # for reactome collection

gsea_results_reac <- GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH", 
  TERM2GENE = dplyr::select(
    mm_reactome_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_reac, x = "NES", showCategory = 30)+ ggtitle("GSEA LFC Reactome")

gsea_results_reac %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_reactome_sets")

```

### Biocarta

No terms enriched

### Wikipathways

```{r, fig.height = 5, fig.width = 8}

mm_wiki_sets <- msigdbr(
  species = "Mus musculus", 
  category = "C2",
  subcategory = "CP:WIKIPATHWAYS") # for Wikipathways collection

gsea_results_wiki <- GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH", 
  TERM2GENE = dplyr::select(
    mm_wiki_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_wiki, x = "NES", showCategory = 30)+ ggtitle("GSEA LFC Wikipathways")

gsea_results_wiki %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_wiki_sets")

```

### KEGG

```{r, fig.height = 5, fig.width = 8}

mm_kegg_sets <- msigdbr(
  species = "Mus musculus", 
  category = "C2",
  subcategory = "CP:KEGG") # for KEGG collection

gsea_results_kegg <- GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH", 
  TERM2GENE = dplyr::select(
    mm_kegg_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_kegg, x = "NES", showCategory = 30)+ ggtitle("GSEA LFC KEGG")

gsea_results_kegg %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_kegg_sets")

```

### TFT (regulatory genes)

No terms enriched

## Pval ranking

!! I mostly did not find enriched terms so I put the acceptance threshold of the enrichment analysis at a pval of 0.05. Meaning this is only for info, not exploitable!

```{r}

ordered_genes_pval <- res_tbl1$padj
names(ordered_genes_pval) <- res_tbl1$gene
ordered_genes_pval <- sort(ordered_genes_pval, decreasing = T)

set.seed(054057)
```

### Gene ontology

```{r}

mm_GO_sets <- msigdbr(
  species = "Mus musculus", 
  category = "C5") # for gene ontology collection

gsea_results_go <- GSEA(
  geneList = ordered_genes_pval, 
  pvalueCutoff = 0.5, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH",
  scoreType = "pos",
  TERM2GENE = dplyr::select(
    mm_GO_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_go, x = "NES", showCategory = 30)+ ggtitle("GSEA pval Gene ontology")

gsea_results_go %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_GO_sets")

```

### Hallmark

No enriched terms under pval of 0.8!

### Reactome

```{r, fig.height = 8, fig.width = 8}

mm_reactome_sets <- msigdbr(
  species = "Mus musculus", 
  category = "C2",
  subcategory = "CP:REACTOME") # for reactome collection

gsea_results_reac <- GSEA(
  geneList = ordered_genes_pval, 
  pvalueCutoff = 0.5, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH", 
  TERM2GENE = dplyr::select(
    mm_reactome_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_reac, x = "NES", showCategory = 30)+ ggtitle("GSEA pval Reactome")

gsea_results_reac %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_reactome_sets")

```

### Biocarta

No term enriched

### Wikipathways

```{r}

mm_wiki_sets <- msigdbr(
  species = "Mus musculus", 
  category = "C2",
  subcategory = "CP:WIKIPATHWAYS") # for Wikipathways collection

gsea_results_wiki <- GSEA(
  geneList = ordered_genes_pval, 
  pvalueCutoff = 0.5, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH", 
  TERM2GENE = dplyr::select(
    mm_wiki_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_wiki, x = "NES", showCategory = 30)+ ggtitle("GSEA pval Wikipathways")

gsea_results_wiki %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_wiki_sets")

```

### KEGG

No terms enriched

### TFT (regulatory genes)

```{r, fig.height = 5, fig.width = 8}

mm_regulatory_sets <- msigdbr(
  species = "Mus musculus", 
  category = "C3",
  subcategory = "TFT:TFT_Legacy") # for regulatory genes collection

gsea_results_tft <- GSEA(
  geneList = ordered_genes_pval, 
  pvalueCutoff = 0.5, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH", 
  TERM2GENE = dplyr::select(
    mm_regulatory_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_tft, x = "NES", showCategory = 30)+ ggtitle("GSEA pval Regulatory genes")

gsea_results_tft %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_regulatory_sets")

```

## Stat ranking

```{r}
ordered_genes_stat <- res_tbl1$stat
names(ordered_genes_stat) <- res_tbl1$gene
ordered_genes_stat <- sort(ordered_genes_stat, decreasing = T)

set.seed(054057)

```

### Gene ontology

```{r}

mm_GO_sets <- msigdbr(
  species = "Mus musculus", 
  category = "C5") # for gene ontology collection

gsea_results_go <- GSEA(
  geneList = ordered_genes_stat, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH", 
  TERM2GENE = dplyr::select(
    mm_GO_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_go, x = "NES", showCategory = 30)+ ggtitle("GSEA Stat Gene ontology")

gsea_results_go %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_GO_sets")

```

### Hallmark

```{r}

mm_hallmark_sets <- msigdbr(
  species = "Mus musculus", 
  category = "H") # for hallmark collection

gsea_results_hall <- GSEA(
  geneList = ordered_genes_stat, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH", 
  TERM2GENE = dplyr::select(
    mm_hallmark_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_hall, x = "NES", showCategory = 30)+ ggtitle("GSEA Stat Hallmarks")

gsea_results_hall %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_hallmark_sets")
```

### Reactome

```{r}

mm_reactome_sets <- msigdbr(
  species = "Mus musculus", 
  category = "C2",
  subcategory = "CP:REACTOME") # for reactome collection

gsea_results_reac <- GSEA(
  geneList = ordered_genes_stat, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH", 
  TERM2GENE = dplyr::select(
    mm_reactome_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_reac, x = "NES", showCategory = 30)+ ggtitle("GSEA Stat Reactome")

gsea_results_reac %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_reactome_sets")

```

### Biocarta

No terms enriched

### Wikipathways

```{r}

mm_wiki_sets <- msigdbr(
  species = "Mus musculus", 
  category = "C2",
  subcategory = "CP:WIKIPATHWAYS") # for Wikipathways collection

gsea_results_wiki <- GSEA(
  geneList = ordered_genes_stat, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH", 
  TERM2GENE = dplyr::select(
    mm_wiki_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_wiki, x = "NES", showCategory = 30)+ ggtitle("GSEA Stat Wikipathways")

gsea_results_wiki %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_wiki_sets")

```

### KEGG

```{r, fig.height = 5, fig.width = 8}

mm_kegg_sets <- msigdbr(
  species = "Mus musculus", 
  category = "C2",
  subcategory = "CP:KEGG") # for KEGG collection

gsea_results_kegg <- GSEA(
  geneList = ordered_genes_stat, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  seed = TRUE, 
  pAdjustMethod = "BH", 
  TERM2GENE = dplyr::select(
    mm_kegg_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results_kegg, x = "NES", showCategory = 30)+ ggtitle("GSEA Stat KEGG")

gsea_results_kegg %>% 
  as_tibble() %>% 
  select(1:9) %>%
  knitr::kable()

rm(list = "mm_kegg_sets")

```

### TFT (regulatory genes)

No terms enriched
